---
title: "Reproducible Research – Course Project 2"
author: "David Levy"
date: "11/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

Title: Your document should have a title that briefly summarizes your data analysis
Synopsis: Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.


## Data Processing

### R Packages Used for Analysis

```{r Calling analysis libraries}

library(tidyverse)
library(ggplot2)

```

### Unzipping and Reading Data

We begin data processing by unzipping and reading our data into memory:

```{r Unzipping and reading data, cache=TRUE}

# store data file name
dataFile  <- 'repdata_data_StormData.csv.bz2'

# download the file if it doesn't exist already
if(!file.exists(dataFile)) {
    download.file('https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2',
                  destfile = dataFile, method = 'curl')
}

# initiate tibble with file contents
bzConnection <- bzfile(dataFile)
stormData <- read.csv(bzConnection)

```

The first thing we can do to try and understand the data set we are working with is to visualize the top and bottom of the tibble we created snd to inquire about the structure of the data frame:

```{r Inspecting the data frame}

# visualize the top of the tibble
head(stormData)

# visualize the bottom of the tibble
tail(stormData)

# summarize the structure of the tibble
str(stormData)
```

There should be a section titled Data Processing which describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the \color{red}{\verb|cache = TRUE|}cache = TRUE option for certain code chunks.


## Results

There should be a section titled Results in which your results are presented.
You may have other sections in your analysis, but Data Processing and Results are required.
The analysis document must have at least one figure containing a plot.
Your analysis must have no more than three figures. Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.
You must show all your code for the work in your analysis document. This may make the document a bit verbose, but that is okay. In general, you should ensure that \color{red}{\verb|echo = TRUE|}echo = TRUE for every code chunk (this is the default setting in knitr).
