---
title: "Comparative Impact of Weather Events on Indicators of Human and Economic Health"
author: "David Levy"
date: "11/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

This analysis makes use of the NOAA Storm Database and attempts to answer some basic questions about severe weather events. More specifically, we are seeking answers to the following:


**Research Questions:**

1. *Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}`EVTYPE` variable) are most harmful with respect to population health?*

2. *Across the United States, which types of events have the greatest economic consequences?*


In order to answer these questions, we begin by importing the NOAA Storm Database and vizualizing its contents. We proceed to outline which variables we intend to utilize and which are not helpful in addressing the research questions posed above. Next, we provide R code demonstrating out data paring and transformation process, and proceed to our analysis and results.

The analysis conducted here is brief, but informative. We group and summarize data by event type with injuries and fatalities as variables of interest in assessing the effect of various types of weather events on population health (Research Question 1). Next, we group and summarize data by event type with property damage and crop damage as variables of interest in assessing the economic consequences of various types of weather events (Research Question 2).

We conclude our analysis and results with graphical representations of the data and a summary of findings that will allow other researchers to develop further research questions related to the effects of weather events on human and economic health.


## Data Processing

#### R Packages Used for Analysis

```{r Calling analysis libraries}

library(tidyverse)
library(ggplot2)

```

#### Unzipping and Reading Data

We begin data processing by unzipping and reading the compressed Storm Data data file into memory:

```{r Unzipping and reading data, cache=TRUE}

# store data file name
dataFile  <- 'repdata_data_StormData.csv.bz2'

# download the file if it doesn't exist already
if(!file.exists(dataFile)) {
    download.file('https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2',
                  destfile = dataFile, method = 'curl')
}

# initiate a connection to the data file
bzConnection <- bzfile(dataFile)

# initiate tibble with file contents
stormData <- read.csv(bzConnection)

```

In order to better understand the data set as it has been provided, we can visualize the top and bottom of our new tibble and get a printout of the structure of the data:

```{r Inspecting the data frame}

# visualize the top of the tibble
head(stormData)

# visualize the bottom of the tibble
tail(stormData)

# summarize the structure of the tibble
str(stormData)

```

#### Data Processing and Transformations

Based on what we observed in the previous section, this data set is bit unwieldy in terms of its width, meaning that we likely have more variables than we need in order to answer the two questions emphasized in the **Synopsis** section. Recall that of particular interest are the `EVTYPE` and its relationship with indicators of (1) population health and (2) economic consequences. Below is a summary of how each variable has been treated in this analysis and a brief rationale for each respective decision:

- `STATE__`   : We will drop this variable. It is redundant and we do not need a numerical representation of the state of storm events for the present analysis.
- `BGN_DATE`  : We will drop this variable. There are no questions pertaining to the temporal aspects of the storm events recorded in this data set.
- `BGN_TIME`  : We will drop this variable. There are no questions pertaining to the temporal aspects of the storm events recorded in this data set.
- `TIME_ZONE` : We will drop this variable. There are no questions pertaining to the temporal aspects of the storm events recorded in this data set.
- `COUNTY`    : We will drop this variable. There are no questions pertaining to spatial aspects of the storm events recorded in this data set.
- `COUNTYNAME`: We will drop this variable. There are no questions pertaining to spatial aspects of the storm events recorded in this data set.
- `STATE`     : We will retain this variable for our analysis. A within-state comparison may be germane to questions of economic impact.
- `EVTYPE`    : We will retain this variable for our analysis. This variable is central to the questions at hand, as both are asked within the framework of of event type. In order to better aggregate data, we will manipulate this variable to be in all capital letters and to omit all vowels, white space and punctuation. This transformation is performed in hopes of avoiding underestimates of injury, fatality, property damage and crop damage totals for various event types.
- `BGN_RANGE` : We will drop this variable. It does not contain information germane to the questions at hand.
- `BGN_AZI`   : We will drop this variable. It does not contain information germane to the questions at hand.
- `BGN_LOCATI`: We will drop this variable. It does not contain information germane to the questions at hand.
- `END_DATE`  : We will drop this variable. There are no questions pertaining to the temporal aspects of the storm events recorded in this data set.
- `END_TIME`  : We will drop this variable. There are no questions pertaining to the temporal aspects of the storm events recorded in this data set.
- `COUNTY_END`: We will drop this variable. It does not contain information germane to the questions at hand.
- `COUNTYENDN`: We will drop this variable. It does not contain information germane to the questions at hand.
- `END_RANGE` : We will drop this variable. It does not contain information germane to the questions at hand.
- `END_AZI`   : We will drop this variable. It does not contain information germane to the questions at hand.
- `END_LOCATI`: We will drop this variable. It does not contain information germane to the questions at hand.
- `LENGTH`    : We will drop this variable. It does not contain information germane to the questions at hand.
- `WIDTH`     : We will drop this variable. It does not contain information germane to the questions at hand.
- `F`         : We will drop this variable. It does not contain information germane to the questions at hand.
- `MAG`       : We will drop this variable. It does not contain information germane to the questions at hand.
- `FATALITIES`: We will retain this variable for our analysis. Fatalities are an important measure of the impact of storms on population health.
- `INJURIES`  : We will retain this variable for our analysis. Injuries are an important measure of the impact of storms on population health.
- `PROPDMG`   : We will retain this variable for our analysis, as it contains a numerical representation (without an order of magnitude) of the property damage caused by the corresponding storm event. Property damage is an important measure of the economic toll taken by storms.
- `PROPDMGEXP`: We will retain this variable for our analysis, as it contains an indication of the order of magnitude of the property damage caused by the corresponding storm event (H,K,M B correspond with hundred, thousand, million and billions respectively). Property damage is an important measure of the economic toll taken by storms.
- `CROPDMG`   : We will retain this variable for our analysis, as it contains a numerical representation (without an order of magnitude) of the crop damage caused by the corresponding storm event. Crop damage is an important measure of the economic toll taken by storms.
- `CROPDMGEXP`: We will retain this variable for our analysis, as it contains an indication of the order of magnitude of the crop damage caused by the corresponding storm event (H,K,M B correspond with hundred, thousand, million and billions respectively). Crop damage is an important measure of the economic toll taken by storms.
- `WFO`       : We will drop this variable. It does not contain information germane to the questions at hand.
- `STATEOFFIC`: We will drop this variable. It does not contain information germane to the questions at hand.
- `ZONENAMES` : We will drop this variable. It does not contain information germane to the questions at hand.
- `LATITUDE`  : We will drop this variable. It does not contain information germane to the questions at hand.
- `LONGITUDE` : We will drop this variable. It does not contain information germane to the questions at hand.
- `LATITUDE_E`: We will drop this variable. It does not contain information germane to the questions at hand.
- `LONGITUDE_`: We will drop this variable. It does not contain information germane to the questions at hand.
- `REMARKS`   : We will drop this variable. It does not contain information germane to the questions at hand.
- `REFNUM`    : We will drop this variable. It does not contain information germane to the questions at hand.

Now that we have summarized how we intend to use (or not use) each variable in the data set, we can transform data appropriately:

```{r Data processing and transformations}

# data transformations and column selections
stormDataFinal <- stormData %>%
    mutate( # remove vowels, spaces, punctuation and terminal "S"s
        event = str_remove_all(str_to_upper(EVTYPE), "(A|E|I|O|U|[:space:]|[:punct:]|(S(?=$)))"), 
        PropertyDamage = ifelse(grepl("H", PROPDMGEXP), PROPDMG * 10^2, # H means hundreds
                         ifelse(grepl("K", PROPDMGEXP), PROPDMG * 10^3, # K means thousands
                         ifelse(grepl("M", PROPDMGEXP), PROPDMG * 10^6, # M means millions
                         ifelse(grepl("B", PROPDMGEXP), PROPDMG * 10^9, # B means billions
                         PROPDMG)))),                                   # otherwise, no transformation
        CropDamage   =   ifelse(grepl("H", CROPDMGEXP), CROPDMG * 10^2, # H means hundreds
                         ifelse(grepl("K", CROPDMGEXP), CROPDMG * 10^3, # K means thousands
                         ifelse(grepl("M", CROPDMGEXP), CROPDMG * 10^6, # M means millions
                         ifelse(grepl("B", CROPDMGEXP), CROPDMG * 10^9, # B means billions
                         CROPDMG))))                                    # otherwise, no transformation
        ) %>%
    select(Event = event, EVTYPE, Fatalities = FATALITIES, Injuries = INJURIES, 
           PropertyDamage, CropDamage) # take only relevant variables

# visualize the new data set
head(stormDataFinal, n=20)
```
Now our data looks usable, so we can proceed to analysis and results in order to answer our research questions.



## Results

#### Population Health Impact

Our first Research Question warrants a closer look at summary statistics on the `fatalities` and `injuries` variables, as they are the two indicators of population health included in the data set. That said, we should point out that there is no use in understanding measures o central tendency, as Research Question 1 asks about the most harmful events across the United States. Thus, we are more interested in aggregate statistics.

A true analysis would require some verified metric to put `fatalities` and `injuries` into comparable units. DALYs might be appropriate, but without a more robust understanding of what qualifies as an "injury" we will simply resort to a sum of the two variables to serve as a proxy for an aggregate measure of population health impact.

```{r Examining summary statistics on population health indicators, message=FALSE}

# group, aggregate, arrange and select the top 5 most harmful weather events 
# to population health
popHealthImpact <- stormDataFinal %>%
    group_by(Event) %>%
    summarise(across(c('Fatalities', 'Injuries'), sum)) %>%
    arrange(desc(Fatalities + Injuries)) %>%
    head(5)

# display the aggregated data
popHealthImpact
```

Observing the console output, it seems that the top five weather events with the highest population health toll in descending order are:

1. Tornado (`TRND`)
2. Excessive Heat (`XCSSVHT`)
3. Thunderstorm Wind (`TSTMWND`)
4. Flood (`FLD`)
5. Lightning (`LGHTNNG`)

We can also display this information graphically:

```{r Plotting population health data}

# gather data for a grouped bar chart
popHealthImpact %>%
    gather('Fatalities', 'Injuries', key = Type, value = Count) %>%
    ggplot(aes(fill=Type, y=Count, x=Event)) +
        geom_bar(position='dodge', stat='identity') +
        theme_bw()
 
```


#### Economic Impact

Our second Research Question can be addressed in much the same way, but instead looking at aggregate statistics on the `propertydamage` and `cropdamage` variables, as they are the two indicators of economic impact that we created through an appropriate transformation of the original data set as described in the previous section. We will sum the aggregate figures in order to get an overall understanding of the economic impact in dollars of each type of weather event:

```{r Examining summary statistics on economic consequences, message=FALSE}

# group, aggregate, arrange and select the top 5 most economically 
# harmful weather events
economicImpact <- stormDataFinal %>%
    group_by(Event) %>%
    summarise(across(c('PropertyDamage', 'CropDamage'), sum, .names = '{.col}')) %>%
    arrange(desc(PropertyDamage + CropDamage)) %>%
    head(5)

# display the aggregated data
economicImpact
```

Observing the output of our code, it seems that the top five most economically harmful weather events in descending order are:

1. Flood (`FLD`)
2. Hurricane/Typhoon (`HRRCNTYPHN`)
3. Tornado (`TRND`)
4. Storm Surge(`STRMSRG`)
5. Hail (`HL`)

We can also display this information graphically:

```{r Plotting economic impact data}

# gather data for a grouped bar chart
economicImpact %>%
    gather('PropertyDamage', 'CropDamage', key = DamageType, value = Count) %>%
    ggplot(aes(fill=DamageType, y=Count, x=Event)) +
        geom_bar(position='dodge', stat='identity') +
        theme_bw()
 
```